NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-12-08 Manuscript/manuscriptPGPE.tex
Update manuscript

@@

Structural details of lipid headgroups are mainly studied using NMR experiments, which
suggest that the glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, 
glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90},
and the headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. 
%Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
%In contrast to PC lipids (see \cite{botan15} and references therein).
Some attempts to resolve conformational ensembles from NMR for PC and PE lipids have been made,
but lesser extend for PG or PS lipids \cite{seelig77c,davis83,Semchyschyn04}.
Classical molecular dynamics simulations could potentially give such ensembles and therefore enable
the detailed studies of lipid headgroup behaviour in complex biomolecular systems, but current
force fields are not accurate enough to reproduce the correct conformational ensembles for PC and PS headgroups \cite{botan15,antila19}.
Several MD simulations of PE and PG lipids have been published especially in the context of modeling
inner membrane of Gram-negative
bacteria \cite{devries04,murzyn05,pedersen06,zhao07,gurtovenko08,zhao08,henin09,kukol09,tsai12,dickson12,venable13,dickson14,berglund15}
\todo{There may be some relevant publication missing from here},
but evaluation of glycerol backbone and headgroup structures against experiments is rare \cite{henin09}.

>>>>

Lipid composition of biomembranes vary greatly between different
organelles and organism, and different lipid types interact with
proteins in many different ways regulating their function \cite{??}.
Lipids may directly interact with proteins by specific binding sites
or indirectly affect protein functions by altering membrane properties
or charge \cite{??}. Specific interactions with certain lipid headgroups
are known to be essential for the function of several proteins \cite{??}.
However, it is not clear if the specific binding is driven by
the differences in acessible conformational states between lipid types or
by specific intermolecular interactions between lipids and proteins.

Differences between preferred conformations of different lipid types
are observed in lipid crystal structures \cite{buldt81,pascher92},
but lipids are not in solid state in biomembranes. 
On the other hand, NMR experiments enable detection of lipid headgroups
in physiological liquid states also in cells \cite{gally81,scherer87,seelig90}.
These studies suggest that the glycerol backbone structures are largely similar irrespectively of the headroup \cite{gally81}, 
glycerol backbone and headgroup structure and behaviour are similar in model membranes and in bacteria \cite{gally81,scherer87,seelig90},
and the headgroup structures are similar in PC, PE and PG lipids, while headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. 
%Extensive discussion about structural details of PE, PG or PS headgroups do not exists (as far as I know), 
%In contrast to PC lipids (see \cite{botan15} and references therein).
%Some attempts to resolve conformational ensembles from NMR for PC and PE lipids have been made,
%but lesser extend for PG or PS lipids \cite{seelig77c,davis83,Semchyschyn04}.
Attempts to resolve the conformational ensembles of lipid headgroup rely mainly on
absolute values of C-H bond order parameters measured using $^2$H NMR~\cite{seelig77c,davis83,Semchyschyn04}.
However, appropriate models to interpret the experimental data have not been available and
also the signs of the order parameters are required to fully capture the details in conformational ensembles  \cite{botan15,ollila16}.
%determine the structural 
%Classical molecular dynamics simulations could potentially give such ensembles and therefore enable
%the detailed studies of lipid headgroup behaviour in complex biomolecular systems, but current
%force fields are not accurate enough to reproduce the correct conformational ensembles for PC and PS headgroups \cite{botan15,antila19}.

@@

%Several MD simulations of PE and PG lipids have been published especially in the context of modeling
%inner membrane of Gram-negative
%bacteria \cite{devries04,murzyn05,pedersen06,zhao07,gurtovenko08,zhao08,henin09,kukol09,tsai12,dickson12,venable13,dickson14,berglund15}
%\todo{There may be some relevant publication missing from here},
%but evaluation of glycerol backbone and headgroup structures against experiments is rare \cite{henin09}.
